### 2022-03-03 omicron-generation.tex
spelling check

There is currently a lack of generation-interval estimates of SARS-CoV-2 varaints, in part, due to difficulties in observing actual infection events.
Instead, many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or testing \citep{abbott2022test} of the infector and the infectee.
However, transmission-interval distributions are subject to dynamical biases and may not be an accurate proxy for the generation-interval distribution.
For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
Therefore, the observed differences (or the lack thereof) in transmission-interval distributions of the invading and resident variants can exaggerate or undermine true differences in their generation-interval distributions, depending on their growth rate differences.

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron varaints from the Netherlands between 13 and 26 December 2021.
The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with and without S gene target failures, but they neglected growth rate differences in their inference.
We first discuss epidemiological context in the Netherlands during the study period and provide corrected estimates for the incubation period.
We combine incubation-period estimates with serial-interval data to infer generation-interval distributions of the Delta and Omicron variants.

@@

To account for growth rate differences between the Delta and Omicron variants, we take the backward incubation-period distributions, estimated by \cite{backer2021omicron}, and calculate the corresponding forward incubation-period distributions by inverting \eref{backward}:
\begin{equation}
f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx)  b_I(x)\dx},
\label{eq:forward}
\end{equation}
where $b_I(\tau)$ is assumed to follow a Weibull distribution.
To further account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
First, we approximate the previously-inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parametrize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified Gamma distributions and estimate the corresponding forward distribution using \eref{forward}.
We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.

@@

For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} previoulsy found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
However, during this period, the number of COVID-19 cases caused by the Delta and Omicron variants showed opposite patterns of epidemic growth/decay (\fref{epidemic}).
Since the number of COVID-19 cases caused by the Delta variant was decreasing, longer incubation periods would have been more likely to be observed.
In contrast, the rapid growth in the number of COVID-19 cases caused by the Omicron variant would have caused shorter incubation periods to be more likely to be observed.
When we account for these growth rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C).
For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$.
The mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.3--4.4 days) to 4.5 days (95\% CI: 3.8--5.6 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.

@@

While we estimate shorter mean forward generation intervals for the Omicron variant across all stratifications (\fref{serial}C), the differences in the mean generation intervals are unclear for between-household pairs.
Consistent with the original study, which also repoted shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
We also estimate a reduction in the mean forward generation intervals from week 50 to week 51.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare.pdf}
\caption{
\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
(A) Observed and fitted forward serial-interval distributions fpr within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
Serial intervals are calculated for infectors who developed symptoms on week 50 (13 and 19 December, 2021).
Points represent the observed data.
Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant. 
Bivariate lognormal distributions with correlation coefficient of 0.75 are used to model relationships between incubation and generation intervals.
(B) Estimated forward generation-interval distributions for within-household transmission pairs in the Netherlands.
(C) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines) stratified by the week of infectors' symptom onset and the type of transmission pairs.
\label{fig:serial}
}
\end{figure}

Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.83 (95\% CI: 1.65--2.05) to 1.17 (95\% CI: 1.00--1.36) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
On the other hand, the reproduction number of the Delta variant decreased from 0.89 (95\% CI: 0.82--0.97) to 0.68 (95\% CI: 0.62--0.74) between December 5, 2021, and January 9, 2022, and increased back up to 0.82 (95\% CI: 0.72--0.93) by January 23, 2022 (\fref{reproduction}A).
We estimate the reproduction number ratios stayed rouhgly constant at around 2.2 (95\% CI: 2.0--2.5) between December 12--26, 2021, and slowly decreased to 1.4 (95\% CI: 1.2--1.7).
However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
In this case, the reproduction ratio decreases from 2.5 (95\% CI: 2.2--2.8) to 1.5 (95\% CI: 1.2--1.8), corresponding to roughly 7--13\% bias.

@@

The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides cruicial information for epidemic control.
A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
Here, we build on our previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
These effects are also pertinent to all epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.

To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant.
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections \citep{park2020inferring,hart2022generation}. 
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structres that involve repeated contacts between the same group of individuals (beacuse only the first infectious contact results in infection).
The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shoter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.

While our study indicate that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how much their infectiousness profiles differ intrinsically.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their intrinsic infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accouting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation interval distribution, which may have a longer mean).

@@

We primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
In doing so, we implicitly assume thta the delay between infection and reports is fixed.
However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
While this limitation does not affect our generation-interval estimates, our inferences of the transmission advantages of the Omicron variant should be interpreted with care.

Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
Our study sythesizes previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters;
however, uncertainty remains in the intrinsic infectiousness profiles of SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.
Future studies extending our framework and combining more detailed epidemiological data will be critical to narrowing down these uncertainties.

>>>>

There is currently a lack of generation-interval estimates of SARS-CoV-2 variants, in part, due to difficulties in observing actual infection events.
Instead, many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or testing \citep{abbott2022test} of the infector and the infectee.
However, transmission-interval distributions are subject to dynamical biases and may not be an accurate proxy for the generation-interval distribution.
For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
Therefore, the observed differences (or the lack thereof) in transmission-interval distributions of the invading and resident variants can exaggerate or undermine true differences in their generation-interval distributions, depending on their growth rate differences.

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with and without S gene target failures, but they neglected growth rate differences in their inference.
We first discuss epidemiological context in the Netherlands during the study period and provide corrected estimates for the incubation period.
We combine incubation-period estimates with serial-interval data to infer generation-interval distributions of the Delta and Omicron variants.

@@

To account for growth rate differences between the Delta and Omicron variants, we take the backward incubation-period distributions, estimated by \cite{backer2021omicron}, and calculate the corresponding forward incubation-period distributions by inverting \eref{backward}:
\begin{equation}
f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx)  b_I(x)\dx},
\label{eq:forward}
\end{equation}
where $b_I(\tau)$ is assumed to follow a Weibull distribution.
To further account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
First, we approximate the previously-inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parameterize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified Gamma distributions and estimate the corresponding forward distribution using \eref{forward}.
We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.

@@

For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} previously found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
However, during this period, the number of COVID-19 cases caused by the Delta and Omicron variants showed opposite patterns of epidemic growth/decay (\fref{epidemic}).
Since the number of COVID-19 cases caused by the Delta variant was decreasing, longer incubation periods would have been more likely to be observed.
In contrast, the rapid growth in the number of COVID-19 cases caused by the Omicron variant would have caused shorter incubation periods to be more likely to be observed.
When we account for these growth rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C).
For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$.
The mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.3--4.4 days) to 4.5 days (95\% CI: 3.8--5.6 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.

@@

While we estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), the differences in the mean generation intervals are unclear for between-household pairs.
Consistent with the original study, which also reported shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
We also estimate a reduction in the mean forward generation intervals from week 50 to week 51.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare.pdf}
\caption{
\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
(A) Observed and fitted forward serial-interval distributions for within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
Serial intervals are calculated for infectors who developed symptoms on week 50 (13 and 19 December, 2021).
Points represent the observed data.
Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant. 
Bivariate lognormal distributions with correlation coefficient of 0.75 are used to model relationships between incubation and generation intervals.
(B) Estimated forward generation-interval distributions for within-household transmission pairs in the Netherlands.
(C) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines) stratified by the week of infectors' symptom onset and the type of transmission pairs.
\label{fig:serial}
}
\end{figure}

Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.83 (95\% CI: 1.65--2.05) to 1.17 (95\% CI: 1.00--1.36) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
On the other hand, the reproduction number of the Delta variant decreased from 0.89 (95\% CI: 0.82--0.97) to 0.68 (95\% CI: 0.62--0.74) between December 5, 2021, and January 9, 2022, and increased back up to 0.82 (95\% CI: 0.72--0.93) by January 23, 2022 (\fref{reproduction}A).
We estimate the reproduction number ratios stayed roughly constant at around 2.2 (95\% CI: 2.0--2.5) between December 12--26, 2021, and slowly decreased to 1.4 (95\% CI: 1.2--1.7).
However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
In this case, the reproduction ratio decreases from 2.5 (95\% CI: 2.2--2.8) to 1.5 (95\% CI: 1.2--1.8), corresponding to roughly 7--13\% bias.

@@

The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides crucial information for epidemic control.
A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
Here, we build on our previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
These effects are also pertinent to all epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.

To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant.
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections \citep{park2020inferring,hart2022generation}. 
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shorter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.

While our study indicate that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how much their infectiousness profiles differ intrinsically.
In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their intrinsic infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation interval distribution, which may have a longer mean).

@@

We primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
In doing so, we implicitly assume that the delay between infection and reports is fixed.
However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
While this limitation does not affect our generation-interval estimates, our inferences of the transmission advantages of the Omicron variant should be interpreted with care.

Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
Our study synthesizes previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters;
however, uncertainty remains in the intrinsic infectiousness profiles of SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.
Future studies extending our framework and combining more detailed epidemiological data will be critical to narrowing down these uncertainties.
